Generating Random Logic Programs Using Constraint Programming

### 2020-01-24 paper/paper.tex
Most of the writing

\documentclass[letterpaper]{article}
\usepackage{uai2019}
\usepackage[margin=1in]{geometry}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}
\usepackage[nounderscore]{syntax}
\usepackage{blkarray}
\usepackage{siunitx}
\usepackage{amsthm}
\usepackage[round]{natbib}
\usepackage[inline]{enumitem}

@@

% Set the typeface to Times Roman
\usepackage{times}

@@

%TODO: define all the relevant terminology from logic and constraint
% programming: predicates ,arities, (logic) variables, constants, clauses, atoms,
% heads, bodies,
% CP: decision variables, constraints.
% TODO: decide between 'argument' and 'gap' and be consistent.
% TODO: need to introduce the idea of predicate independence

If a predicate has arity $n$ and it is as of yet undecided what $n$ terms will
fill those spots, we will say that the predicate has $n$ \emph{gaps}. We say
that a constraint variable is \emph{(fully) determined} if its domain (at the
given moment in the execution) has exactly one value.

@@

\begin{definition}
  The \emph{head} of a clause is composed of
  \begin{itemize}
  \item a $\variable{predicate} \in \predicates \cup \{ \Box \}$.
  \item and $\arrayd{arguments}{\maxArity{}}{\constants{} \cup \variables{}}
    \cup \{ \Box \}$
  \end{itemize}
\end{definition}
The reason why $\Box$ must be a separate value will become clear in
\cref{sec:variable_symmetry}.

@@

@@

\begin{constraint}
  Each predicate gets at least one clause. Let
  \[
    P = \{ h.\variable{predicate} \mid h \in \variable{heads} \}.
  \]
  Then
  \[
    \variable{nValues}(P) =
    \begin{cases}
      |\predicates{}| + 1 & \text{if } \variable{count}(\Box, P) > 0 \\
      |\predicates{}| & \text{otherwise.}
    \end{cases}
  \]
  Here, $\variable{nValues}(P)$ counts the number of unique values in $P$.
\end{constraint}

@@

\begin{definition}
  The body of a clause is defined by:
  \begin{itemize}
  \item $\arrayd{structure}{\maxNumNodes{}}{[0, \maxNumNodes{} - 1]}$ such
    that:
    \begin{itemize}
    \item $\variable{structure}[i] = i$: the $i$-th node is a root.
    \item $\variable{structure}[i] = j$: the $i$-th node's parent is node $j$.
    \end{itemize}
  \item $\arrayt{values}{\maxNumNodes{}}{Node}$.
  \end{itemize}
\end{definition}

@@

\begin{example} \label{example:formula}
  Let $\maxNumNodes{} = 8$. Then $\neg\mathsf{P}(X) \lor (\mathsf{Q}(X)
  \land \mathsf{P}(X))$ corresponds to the tree in \cref{fig:example_tree} and
  can be encoded as:
  \begin{alignat*}{8}
    \variable{structure} &= [0, &&0, &&0, &&1, &&2, &&2, &&6, &&7],\\
    \variable{values} &= [{\lor}, &&{\neg}, &&{\land}, \mathsf{P}(&&X), \mathsf{Q}(&&X), \mathsf{P}(&&X), &&\top, &&\top],\\
    \variable{numNodes} &= 6, && && && && && && &&\\
    \variable{numTrees} &= 3. && && && && && && &&
  \end{alignat*}
  In the rest of this section, we will describe how the elements of
  $\variable{values}$ are encoded and list a series of constraints that make
  this representation unique.
\end{example}

@@

We can use \cref{constr:arity} again to disable extra arguments.

\begin{example}
  Let $\maxArity{} = 2$, $\predicates{} = [\mathsf{P}, \dots]$, $\arities{}
  = [1, \dots]$, and $X \in \variables{}$. Then the node representing atom
  $\mathsf{P}(X)$ has:
  \begin{align*}
    \variable{name} &= \mathsf{P},\\
    \variable{arguments} &= [X, \Box],\\
    \variable{arity} &= 1.
  \end{align*}
\end{example}

\subsection{CONSTRAINTS}

\begin{definition}
  Let $\variable{numTrees} \in \{ 1, \dots, \maxNumNodes{} \}$ be the number of
  trees in $\variable{structure}$, defined using the $\variable{tree}$
  constraint \citep{DBLP:conf/cp/FagesL11}.
\end{definition}

@@

@@

\begin{constraint} \label{constraint:node_types}
  For $i = 0, \dots, \maxNumNodes{} - 1$, let $C_i$ be the number of times $i$
  appears in the \variable{struture} array with index greater than $i$. Then
  \begin{align*}
    C_i = 0 &\iff \variable{values}[i].\variable{name} \in \predicates{} \cup \{ \top \},\\
    C_i = 1 &\iff \variable{values}[i].\variable{name} = \neg,\\
    C_i > 1 &\iff \variable{values}[i].\variable{name} \in \{ \land, \lor \}.
  \end{align*}
\end{constraint}
Each of the three equivalences in \cref{constraint:node_types} corresponds to
node $i$ having no children, one child, and multiple children, respectively.

\begin{constraint}
  Only root nodes can have $\top$ as the value. For $i = 0, \dots,
  \maxNumNodes{} - 1$,
  \[
    \variable{structure}[i] \ne i \implies
    \variable{values}[i].\variable{name} \ne \top.
  \]
\end{constraint}

A way to `disable' a clause:
\begin{constraint}
  For $i = 0, \dots, \maxNumClauses{} - 1$, if
  $\variable{heads}[i].\variable{predicate} = \Box$, then
  \[
    \variable{bodies}[i].\variable{numNodes} = 1,
  \]
  and
  \[
    \variable{bodies}[i].\variable{values}[0].\variable{name} =
    \top.
  \]
\end{constraint}

@@

\begin{definition}
  Let $N = \maxArity{} \times (\maxNumNodes{} + 1)$. Let
  $\variable{terms}[N] \in \constants{} \cup \variables{} \cup \{ \Box
  \}$ be a flattened array of all gaps in a particular clause.

@@

The number of ways to fill $n$ gaps with terms can be expressed as
\begin{align*}
  P(n) = |\constants{}|^n + &\sum_{\substack{1 \le k \le |\variables{}|, \\ 0 =
      s_0 < s_1 < \dots < s_k < s_{k+1} = n+1}}\\
  &\prod_{i=0}^k (|\constants{}| + i)^{s_{i+1} - s_i - 1}.
\end{align*}
The first term is the number of ways to fill all $n$ gaps with
constants. The parameter $k$ is the number of variables used in the clause, and
$s_1, \dots, s_k$ mark the first occurrence of each variable. For each gap
between any two introductions (or before the first introduction, or after the last
introduction), we have $s_{i+1}-s_i-1$ spaces to be filled with any of the
$\constants{}$ constants or any of the $i$ already-introduced variables.

Let us order the elements of $\predicates{}$, and let $a_i$ be the arity of the
$i$-th predicate. The number of programs is then:
\[
  \sum_{\substack{ \sum_{i=1}^{|\predicates{}|} h_i = n,\\
      |\predicates{}| \le n \le \maxNumClauses,\\
      h_i \ge 1 \text{ for all } i}} \prod_{i=1}^{|\predicates{}|}
  \multiset{\sum_{a=0}^{\maxArity{} \times \maxNumNodes{}} C(a) P(a+a_i)}{h_i},
\]
where
\[
  \multiset{n}{k} = \binom{n+k-1}{k}
\]
counts the number of ways to select $k$ out of $n$ items with repetition (and
without ordering). Here, we sum over all possible ways to distribute
$|\predicates{}| \le n \le \maxNumClauses{}$ clauses among $|\predicates{}|$
predicates so that each predicate gets at least one clause. For each predicate,
we can then count the number of ways to select its clauses out of all possible
clauses. The number of possible clauses can be computed by considering each
possible arity $a$, and multiplying the number of `unfinished' clauses $C(a)$ by
the number of ways to fill the $a+a_i$ gaps in the body and the head of the
clause with terms.

@@

\begin{definition}[Adjacency matrix representation] \label{def:adjacency_matrix}
  An $|\predicates{}| \times |\predicates{}|$ adjacency matrix $\mathbf{A}$ with
  $\{ 0, 1 \}$ as its domain is defined by stating that $\mathbf{A}[i][j] = 0$
  if and only if, for all $k \in \{ 0, \dots, \maxNumClauses{} - 1 \}$, either
$\variable{heads}[k].\variable{predicate} \ne j$ or $i \not\in \{
a.\variable{name} \mid a \in \variable{bodies}[k].\variable{values} \}$.
\end{definition}

Given an undetermined model, we can classify all dependencies of a predicate
$\mathsf{P}$ into three categories based on how many of the edges on the path
from the dependency to $\mathsf{P}$ are undetermined. In the case of zero, we
call the dependency \emph{determined}. In the case of one, we call it
\emph{almost determined}. Otherwise, it is \emph{undetermined}. In the context
of propagation and entailment algorithms, we define a \emph{dependency} as the
sum type:
\begin{grammar}
  <dependency> ::= $\Determined(p)$ | $\Undetermined(p)$ | $\AlmostDetermined(p,
  s, t)$
\end{grammar}
where $p \in \predicates{}$ is the name of the predicate which is the
dependency of $\mathsf{P}$, and---in the case of $\AlmostDetermined$---$(s, t)
\in \predicates{}^2$ is the one undetermined edge in $\mathbf{A}$ that prevents
the dependency from being determined. For a dependency $d$---regardless of its
exact type---we will refer to its predicate $p$ as $d.\variable{predicate}$. In
describing the algorithms, we will use an underscore to replace any of $p$, $s$,
$t$ in situations where the name is unimportant.

\begin{algorithm}
  \SetKwFunction{getDependencies}{deps}
  \SetKwFunction{fail}{fail}
  \SetKwFunction{removeValue}{removeValue}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \KwData{predicates $p_1$, $p_2$; adjacency matrix $\mathbf{A}$}
  \For{$(d_1, d_2) \in \getDependencies{$p_1$, 0} \times \getDependencies{$p_2$,
      0}$ such that $d_1.\predicate = d_2.\predicate$}{
    \lIf{$d_1$ {\bf is} $\Determined(\_)$ {\bf and} $d_2$ {\bf is}
      $\Determined(\_)$}{\fail{}}
    \If{$d_1$ {\bf is} $\Determined(\_)$ {\bf and} $d_2$ {\bf is}
      $\AlmostDetermined(\_, s, t)$ {\bf or} \hspace{50pt} $d_2$ {\bf is}
      $\Determined(\_)$ {\bf and} $d_1$ {\bf is} $\AlmostDetermined(\_, s, t)$}{
      $\mathbf{A}[s][t]$.\removeValue{$1$}\;
    }
  }
  \caption{Propagation for independence}
  \label{alg:independence_propagation}
\end{algorithm}
\begin{algorithm}
  \SetKwFunction{getDependencies}{deps}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwData{predicate}{predicate}
  \KwData{predicates $p_1$, $p_2$}
  \nosemic $D \gets \{ (d_1, d_2) \in \getDependencies{$p_1$, 1} \times \getDependencies{$p_2$, 1}$\;
  \hspace{27pt}\dosemic $\mid d_1.\predicate = d_2.\predicate \}$\;
  \lIf{$D = \emptyset$}{\Return{\textsc{true}}}
  \lIf{$\exists (\Determined \_, \Determined \_) \in D$}{\Return{\textsc{false}}}
  \Return{\textsc{undefined}}\;
  \caption{Entailment for independence}
  \label{alg:independence_entailment}
\end{algorithm}
\begin{algorithm}
  \SetKwData{edgeExists}{edge}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \SetKwData{all}{allDependencies}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwFunction{getDependencies}{deps}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{adjacency matrix $\mathbf{A}$}
  \Fn{\getDependencies{$p$, \all}} {
    $D \gets \{ \Determined(p) \}$\;
    \While{\logical{true}}{
      $D' \gets \emptyset$\;
      \For{$d \in D$ {\bf and} $q \in \predicates{}$}{
        $\edgeExists \gets \mathbf{A}[q][d.\predicate] = \{ 1 \}$\;
        \uIf{$\edgeExists$ {\bf and} $d$ {\bf is} $\Determined(\_)$}{
          $D' \gets D' \cup \{ \Determined(q) \}$
        }
        \uElseIf{$\edgeExists$ {\bf and} $d$ {\bf is} $\AlmostDetermined(\_, s,
          t)$}{
          $D' \gets D' \cup \{ \AlmostDetermined(q, s, t) \}$\;
        }
        \uElseIf{$|\mathbf{A}[q][d.\predicate]| > 1$ {\bf and} $d$~{\bf is}~$\Determined(r)$}{
          $D' \gets D' \cup \{ \AlmostDetermined(q, q, r) \}$\;
        }
        \ElseIf{$|\mathbf{A}[q][d.\predicate]| > 1$ {\bf and} \all}{
          $D' \gets D' \cup \{ \Undetermined(q) \}$\;
        }
      }
      \lIf{$D' = D$}{\Return{$D$}}
      $D \gets D'$\;
    }
  }
  \caption{Dependencies of a predicate}
  \label{alg:dependencies}
\end{algorithm}
\begin{figure}
  \centering
  \[
    \begin{blockarray}{ccccccc}
      \begin{block}{ccc(cccc)}
        \mathsf{father} & & & 0 & 0 & 0 & 0 \\
        \mathsf{mother} & & & 1 & 0 & 0 & 0 \\
        \mathsf{parent} & & & 1 & \fbox{\{ 0, 1 \}} & \{ 0, 1 \} & \{ 0, 1 \} \\
        \mathsf{sibling} & & & 0 & 0 & 0 & 0 \\
      \end{block}
    \end{blockarray}
  \]
  \caption{The adjacency matrix defined using \cref{def:adjacency_matrix} for
    \cref{example:independence}}
  \label{fig:dependencies_matrix}
\end{figure}
\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
    \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
    \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
    \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
    \draw[-{Stealth}] (parent) edge (father);
    \draw[-{Stealth}] (mother) edge (father);
    \draw[-{Stealth},dashed,darkgray] (parent) edge (sibling);
    \draw[-{Stealth},dashed,darkgray] (parent) edge (mother);
  \end{tikzpicture}
  \caption{The predicate dependency graph that corresponds to
    \cref{fig:dependencies_matrix}. Dashed edges are undetermined---they may
    or may not exist.}
  \label{fig:dependencies2}
\end{figure}

% TODO: start here
% TODO: make sure my definitions and examples are consistent with loops
% TODO: Discuss that the algorithms work in case there are multiple 'paths'
% without any issues.

Note that if there are multiple paths in the dependency graph from $\mathsf{Q}$
to $\mathsf{P}$, when constructing $D_{\mathsf{P}}$, \cref{alg:dependencies}
could include $\mathsf{Q}$ once for each possible type ($\Determined$,
$\Undetermined$, and $\AlmostDetermined$). However,
\cref{alg:independence_propagation,alg:independence_entailment}

@@

\section{NEGATIVE CYCLES} \label{sec:cycles}
% TODO: NB: the set of decision variables for this constraint must exclude all
% arguments.
% TODO: My paper has two slightly different graphs with the same name.

>>>>

\documentclass[letterpaper]{article}
\usepackage{uai2020}
\usepackage[margin=1in]{geometry}
\usepackage{times}

\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}
\usepackage[nounderscore]{syntax}
\usepackage{blkarray}
\usepackage{siunitx}
\usepackage{amsthm}
\usepackage[round]{natbib}
\usepackage[inline]{enumitem}

@@

@@

%TODO: define all the relevant terminology from logic and constraint
% programming: predicates ,arities, (logic) variables, constants, clauses, atoms,
% heads, bodies,
% CP: decision variables, constraints.
% TODO: need to introduce the idea of predicate independence

We say that a constraint variable is \emph{(fully) determined} if its domain (at
the given moment in the execution) has exactly one value.

@@

\begin{definition}
  The \emph{head} of a clause is composed of a $\variable{predicate} \in
  \predicates \cup \{ \Box \}$, and
  $\arrayd{arguments}{\maxArity{}}{\constants{} \cup \variables{}} \cup \{ \Box
  \}$.
\end{definition}
Here, we use $\Box$ to denote either a disabled clause that we choose not to use
or disabled arguments if the arity of the $\variable{predicate}$ is less than
$\maxArity{}$. The reason why we need a separate value for the latter (i.e., why
it is not enough to fix disabled arguments to a single already-existing value)
will become clear in \cref{sec:variable_symmetry}.

@@

Having defined arity, we can now fix the superfluous arguments:

@@

We can also add a constraint that each predicate $\mathsf{P} \in \predicates{}$
should have at least one clause with $\mathsf{P}$ at its head:

\begin{constraint}
Let
  \[
    P = \{ h.\variable{predicate} \mid h \in \variable{heads} \}.
  \]
  Then
  \[
    \variable{nValues}(P) =
    \begin{cases}
      |\predicates{}| + 1 & \text{if } \variable{count}(\Box, P) > 0 \\
      |\predicates{}| & \text{otherwise.}
    \end{cases}
  \]
\end{constraint}

Here, $\variable{nValues}(P)$ counts the number of unique values in $P$.

@@

As was briefly mentioned before, the body of a clause is represented by a tree.

\begin{definition}
  The \emph{body} of a clause has two parts. First, we have the
  $\arrayd{structure}{\maxNumNodes{}}{[0, \maxNumNodes{} - 1]}$ array that
  encodes the structure of the tree using the following two rules:
  $\variable{structure}[i] = i$ means that the $i$-th node is a root, and
  $\variable{structure}[i] = j$ (for $j \ne i$) means that the $i$-th node's
  parent is node $j$. The second part is the array
  $\arrayt{values}{\maxNumNodes{}}{Node}$ such that $\variable{values}[i]$ holds
  the value of the $i$-th node.
\end{definition}

We can use the $\variable{tree}$ constraint \citep{DBLP:conf/cp/FagesL11} to
forbid cycles in the $\variable{structure}$ array and simultaneously define
$\variable{numTrees} \in \{ 1, \dots, \maxNumNodes{} \}$ to count the number of
trees. We will view the tree rooted at the zeroth node as the main tree and
restrict all other trees to single nodes. For this to work, we need to make sure
that the zeroth node is indeed a root:

@@

\begin{example} \label{example:formula}
  Let $\maxNumNodes{} = 8$. Then
  \[
    \neg\mathsf{P}(X) \lor (\mathsf{Q}(X) \land \mathsf{P}(X))
  \]
  corresponds to the tree in \cref{fig:example_tree} and can be encoded as:
  \begin{alignat*}{8}
    \variable{structure} &= [0, &&0, &&0, &&1, &&2, &&2, &&6, &&7],\\
    \variable{values} &= [{\lor}, &&{\neg}, &&{\land}, \mathsf{P}(&&X), \mathsf{Q}(&&X), \mathsf{P}(&&X), &&\top, &&\top],\\
    \variable{numNodes} &= 6, && && && && && && &&\\
    \variable{numTrees} &= 3. && && && && && && &&
  \end{alignat*}
\end{example}

Here, $\top$ is the value we use for the remaining one-node trees. The
elements of the $\variable{values}$ array are \emph{nodes}.

@@

Furthermore, we can use \cref{constr:arity} again to disable the extra
arguments.

\begin{example}
  Let $\maxArity{} = 2$, $X \in \variables{}$, and let $\mathsf{P}$ be a
  predicate with arity 1. Then the node representing atom $\mathsf{P}(X)$ has:
  \begin{align*}
    \variable{name} &= \mathsf{P},\\
    \variable{arguments} &= [X, \Box],\\
    \variable{arity} &= 1.
  \end{align*}
\end{example}

It remains to constrain the forest represented by the $\variable{structure}$
array together with its $\variable{values}$ to eliminate unnecessary symmetries
and adhere to our desired format. First, we can recognise that the order of the
elements in the $\variable{structure}$ array does not matter, i.e., the
structure is only defined by how the elements link to each other, so we can add
a constraint saying that:

@@

Next, since we already have a variable that counts the number of nodes in the
main tree, we can fix the structure and the values of the remaining trees to
some constant values:

@@

The second part of this constraint states that every node in the main tree
except the zeroth node cannot be a root and must have its parent located to
the left of itself. Next, we classify all nodes into three classes: predicate
(or empty) nodes, negation nodes, and conjunction/disjunction nodes based on the
number of children (zero, one, and two, respectively).

\begin{constraint} \label{constraint:node_types}
  For $i = 0, \dots, \maxNumNodes{} - 1$, let $C_i$ be the number of times $i$
  appears in the \variable{struture} array with index greater than $i$. Then
  \begin{align*}
    C_i = 0 &\iff \variable{values}[i].\variable{name} \in \predicates{} \cup \{ \top \},\\
    C_i = 1 &\iff \variable{values}[i].\variable{name} = \neg,\\
    C_i > 1 &\iff \variable{values}[i].\variable{name} \in \{ \land, \lor \}.
  \end{align*}
\end{constraint}

The value $\top$ serves a twofold purpose: it is used as the fixed value for
nodes outside the main tree, and, when located at the zeroth node, it can
represent a clause with no body. Thus, we can say that only root nodes can have
$\top$ as the value:

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \[
    \variable{structure}[i] \ne i \implies
    \variable{values}[i].\variable{name} \ne \top.
  \]
\end{constraint}

Finally, we add a way to disable a clause by setting its head predicate to
$\Box$:

\begin{constraint}
  For $i = 0, \dots, \maxNumClauses{} - 1$, if
  $\variable{heads}[i].\variable{predicate} = \Box$, then
  \[
    \variable{bodies}[i].\variable{numNodes} = 1,
  \]
  and
  \[
    \variable{bodies}[i].\variable{values}[0].\variable{name} =
    \top.
  \]
\end{constraint}

@@

\begin{definition}
  Let $N = \maxArity{} \times (\maxNumNodes{} + 1)$. Let
  $\variable{terms}[N] \in \constants{} \cup \variables{} \cup \{ \Box
  \}$ be a flattened array of all arguments in a particular clause.

@@

The number of ways to select $n$ terms is
\begin{align*}
  P(n) = |\constants{}|^n + &\sum_{\substack{1 \le k \le |\variables{}|, \\ 0 =
      s_0 < s_1 < \dots < s_k < s_{k+1} = n+1}}\\
  &\prod_{i=0}^k (|\constants{}| + i)^{s_{i+1} - s_i - 1}.
\end{align*}
The first term is the number of ways select $n$ constants. The parameter $k$ is
the number of variables used in the clause, and $s_1, \dots, s_k$ mark the first
occurrence of each variable. For each gap between any two introductions (or
before the first introduction, or after the last introduction), we have
$s_{i+1}-s_i-1$ spaces to be filled with any of the $|\constants{}|$ constants
or any of the $i$ already-introduced variables.

Let us order the elements of $\predicates{}$, and let $a_i$ be the arity of the
$i$-th predicate. The number of programs is then:
\[
  \sum_{\substack{ \sum_{i=1}^{|\predicates{}|} h_i = n,\\
      |\predicates{}| \le n \le \maxNumClauses,\\
      h_i \ge 1 \text{ for all } i}} \prod_{i=1}^{|\predicates{}|}
  \multiset{\sum_{a=0}^{\maxArity{} \times \maxNumNodes{}} C(a) P(a+a_i)}{h_i},
\]
where
\[
  \multiset{n}{k} = \binom{n+k-1}{k}
\]
counts the number of ways to select $k$ out of $n$ items with repetition (and
without ordering). Here, we sum over all possible ways to distribute
$|\predicates{}| \le n \le \maxNumClauses{}$ clauses among $|\predicates{}|$
predicates so that each predicate gets at least one clause. For each predicate,
we can then count the number of ways to select its clauses out of all possible
clauses. The number of possible clauses can be computed by considering each
possible arity $a$, and multiplying the number of `unfinished' clauses $C(a)$ by
the number of ways to select the required $a+a_i$ terms in the body and the head
of the clause.

@@

We can now add a constraint to define an adjacency matrix for the predicate
dependency graph but without positivity/negativity.
\begin{definition} \label{def:adjacency_matrix}
  An $|\predicates{}| \times |\predicates{}|$ adjacency matrix $\mathbf{A}$ with
  $\{ 0, 1 \}$ as its domain is defined by stating that $\mathbf{A}[i][j] = 0$
  if and only if, for all $k \in \{ 0, \dots, \maxNumClauses{} - 1 \}$, either
  \[
    \variable{heads}[k].\variable{predicate} \ne j
  \]
  or
  \[
    i \not\in \{a.\variable{name} \mid a \in
    \variable{bodies}[k].\variable{values} \}.
  \]
\end{definition}

Given an undetermined model, we can classify all dependencies of a predicate
$\mathsf{P}$ into three categories based on how many of the edges on the path
from the dependency to $\mathsf{P}$ are undetermined. In the case of zero, we
call the dependency \emph{determined}. In the case of one, we call it
\emph{almost determined}. Otherwise, it is \emph{undetermined}. In the context
of propagation and entailment algorithms, we define a \emph{dependency} as the
sum type:
\begin{grammar}
  <dependency> ::= $\Determined(p)$ | $\Undetermined(p)$ | $\AlmostDetermined(p,
  s, t)$
\end{grammar}
where each alternative represents a determined, undetermined, and almost
determined dependency, respectively. Here, $p \in \predicates{}$ is the name of
the predicate which is the dependency of $\mathsf{P}$, and---in the case of
$\AlmostDetermined$---$(s, t) \in \predicates{}^2$ is the one undetermined edge
in $\mathbf{A}$ that prevents the dependency from being determined. For a
dependency $d$---regardless of its exact type---we will refer to its predicate
$p$ as $d.\mathsf{predicate}$. In describing the algorithms, we will use an
underscore to replace any of $p$, $s$, $t$ in situations where the name is
unimportant.

\begin{algorithm}
  \SetKwFunction{getDependencies}{deps}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwData{predicate}{predicate}
  \KwData{predicates $p_1$, $p_2$}
  \nosemic $D \gets \{ (d_1, d_2) \in \getDependencies{$p_1$, 1} \times \getDependencies{$p_2$, 1}$\;
  \hspace{27pt}\dosemic $\mid d_1.\predicate = d_2.\predicate \}$\;
  \lIf{$D = \emptyset$}{\Return{\textsc{true}}}
  \lIf{$\exists (\Determined \_, \Determined \_) \in D$}{\Return{\textsc{false}}}
  \Return{\textsc{undefined}}\;
  \caption{Entailment for independence}
  \label{alg:independence_entailment}
\end{algorithm}

Each entailment algorithm returns one out of three different values:
\textsc{true} if the constraint is guaranteed to hold, \textsc{false} if the
constraint is violated, and \textsc{undefined} if whether the constraint will be
satisfied or not depends on the future decisions made by the solver.
\cref{alg:independence_entailment} outlines a simple entailment algorithm for
the independence of two predicates $p_1$ and $p_2$. First, we separately
calculate all dependencies of $p_1$ and $p_2$, and look at the set $D$ of
dependencies that $p_1$ and $p_2$ have in common. If there are none, then the
predicates are clearly independent. If they have a dependency in common that is
already fully determined ($\Determined$) for both predicates, then they cannot be
independent. Otherwise, we return \textsc{undefined}.

\begin{algorithm}
  \LinesNumbered
  \SetKwFunction{getDependencies}{deps}
  \SetKwFunction{fail}{fail}
  \SetKwFunction{removeValue}{removeValue}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \KwData{predicates $p_1$, $p_2$; adjacency matrix $\mathbf{A}$}
  \For{$(d_1, d_2) \in \getDependencies{$p_1$, 0} \times \getDependencies{$p_2$,
      0}$ such that $d_1.\predicate = d_2.\predicate$}{
    \lIf{$d_1$ {\bf is} $\Determined(\_)$ {\bf and} $d_2$ {\bf is}
      $\Determined(\_)$}{\fail{}} \label{line:fail}
    \If{$d_1$ {\bf is} $\Determined(\_)$ {\bf and} $d_2$ {\bf is}
      $\AlmostDetermined(\_, s, t)$ {\bf or} \hspace{50pt} $d_2$ {\bf is}
      $\Determined(\_)$ {\bf and} $d_1$ {\bf is} $\AlmostDetermined(\_, s,
      t)$ \label{line:propagation_condition}}{
      $\mathbf{A}[s][t]$.\removeValue{$1$}\; \label{line:propagation_result}
    }
  }
  \caption{Propagation for independence}
  \label{alg:independence_propagation}
\end{algorithm}

Propagation algorithms have two goals: causing a contradiction (failing) in
situations where the corresponding entailment algorithm would return
\textsc{false}, and eliminating values from domains of variables that are
guaranteed to cause a contradiction. \cref{alg:independence_propagation} does
the former on \cref{line:fail}. Furthermore, for any dependency shared between
predicates $p_1$ and $p_2$, if it is determined ($\Determined$) for one
predicate and almost determined ($\AlmostDetermined$) for another, then the edge
that prevents the $\AlmostDetermined$ from becoming a $\Determined$ cannot
exist--\cref{line:propagation_condition,line:propagation_result} take care of
that.

\begin{algorithm}
  \SetKwData{edgeExists}{edge}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \SetKwData{all}{allDependencies}
  \SetKwFunction{getDependencies}{deps}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{adjacency matrix $\mathbf{A}$}
  \Fn{\getDependencies{$p$, \all}} {
    $D \gets \{ \Determined(p) \}$\;
    \While{\logical{true}}{
      $D' \gets \emptyset$\;
      \For{$d \in D$ {\bf and} $q \in \predicates{}$}{
        $\edgeExists \gets \mathbf{A}[q][d.\predicate] = \{ 1 \}$\;
        \uIf{$\edgeExists$ {\bf and} $d$ {\bf is} $\Determined(\_)$}{
          $D' \gets D' \cup \{ \Determined(q) \}$
        }
        \uElseIf{$\edgeExists$ {\bf and} $d$ {\bf is} $\AlmostDetermined(\_, s,
          t)$}{
          $D' \gets D' \cup \{ \AlmostDetermined(q, s, t) \}$\;
        }
        \uElseIf{$|\mathbf{A}[q][d.\predicate]| > 1$ {\bf and} $d$~{\bf is}~$\Determined(r)$}{
          $D' \gets D' \cup \{ \AlmostDetermined(q, q, r) \}$\;
        }
        \ElseIf{$|\mathbf{A}[q][d.\predicate]| > 1$ {\bf and} \all}{
          $D' \gets D' \cup \{ \Undetermined(q) \}$\;
        }
      }
      \lIf{$D' = D$}{\Return{$D$}}
      $D \gets D'$\;
    }
  }
  \caption{Dependencies of a predicate}
  \label{alg:dependencies}
\end{algorithm}

The function $\getDependencies$ in \cref{alg:dependencies} calculates $D_p$ for
any predicate $p$. It has two versions: $\getDependencies(p, 1)$ returns all
dependencies, while $\getDependencies(p, 0)$ returns only determined and
almost-determined dependencies. It starts by establishing the predicate $p$
itself as a dependence and continues to add dependencies of dependencies until
the set $D$ stabilises. For each dependency $d \in D$, we look at the in-links
of $d$ in the predicate dependency graph. If the edge from some predicate $q$ to
$d.\mathsf{predicate}$ is fully determined and $d$ is determined, then $q$ is
another determined dependency of $p$. If the edge is determined but $d$ is
almost determined, then $q$ is an almost-determined dependency. The same outcome
applies if $d$ is fully determined but the edge is undetermined. Finally, if we
are interested in collecting all dependencies regardless of their status, then
$q$ is a dependency of $p$ as long as the edge from $q$ to
$d.\mathsf{predicate}$ is possible. Note that if there are multiple paths in the
dependency graph from $q$ to $p$, \cref{alg:dependencies} could include $q$ once
for each possible type ($\Determined$, $\Undetermined$, and
$\AlmostDetermined$), but
\cref{alg:independence_propagation,alg:independence_entailment} would still work
as intended.

\begin{figure}
  \centering
  \[
    \begin{blockarray}{ccccc}
      \begin{block}{c(cccc)}
        \mathsf{father} & 0 & 0 & 0 & 0 \\
        \mathsf{mother} & 1 & 0 & 0 & 0 \\
        \mathsf{parent} & 1 & \fbox{\{ 0, 1 \}} & \{ 0, 1 \} & \{ 0, 1 \} \\
        \mathsf{sibling} & 0 & 0 & 0 & 0 \\
      \end{block}
    \end{blockarray}
  \]
  \caption{The adjacency matrix defined using \cref{def:adjacency_matrix} for
    \cref{example:independence}}
  \label{fig:dependencies_matrix}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
    \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
    \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
    \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
    \draw[-{Stealth}] (parent) edge (father);
    \draw[-{Stealth}] (mother) edge (father);
    \draw[-{Stealth},dashed,darkgray] (parent) edge (sibling);
    \draw[-{Stealth},dashed,darkgray] (parent) edge (mother);
  \end{tikzpicture}
  \caption{The predicate dependency graph that corresponds to
    \cref{fig:dependencies_matrix}. Dashed edges are undetermined---they may
    or may not exist.}
  \label{fig:dependencies2}
\end{figure}

@@

\section{NEGATIVE CYCLES} \label{sec:cycles}
% TODO: My paper has two slightly different graphs with the same name.
% TODO: NB: the set of decision variables for this constraint must exclude all
% arguments.
% TODO: maybe move some of the definition 10 here.

Having no negative cycles in the predicate dependency graph is a requirement for
probabilistic logic programming language ProbLog \citep{kimmig2009trading},
although it has been shown how the requirement can be alleviated by introducing
negative probabilities \citep{DBLP:journals/ijar/BuchmanP17}. Ideally, we would
like to design a constraint for that similar to the constraint for independence
in the previous section. However, the difficulty with creating a propagation
algorithm for negative cycles is that there seems to be no good way to extend
\cref{def:adjacency_matrix} so that the adjacency matrix captures
positivity/negativity. Thus, we settle for an entailment algorithm with no
propagation.
